Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-08 paper/paper_draft.qmd
Updates to messaging of depth-connectivity interaction.

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral
    cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    reefs were analysed for reef coral cover time series and their ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour at 15m depth. These findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While reefs experience similar
magnitude DHWs based on SST, the proportion of a reefs' coral population impacted by heat
stress events depends on reef depth. Our results suggest that for non-mesophotic reef
ecosystems (i.e., \< 30m), depth contributes greatly to reef resilience and maintenance
positive carbonate budgets under future conditions. Movement of shallow reefs towards
negative carbonate budgets in the future may contribute to phase shifts in reef dominant
CaCO~3~ production. Shifting from coral driven to encrusting algae driven production, or to
bioerosion dominated systems with rapid erosion of physical reef frameworks [@dudgeon2010;
@perry2008]. Reefs at increased depths may be able to maintain coral driven CaCO~3~
production and accumulation into the future for longer. Spatially these results indicate
that regions of positive carbonate budget maintenance, under ACCESS-ESM1-5, and high depth
levels were concentrated in the Far Northern, offshore Townsville/Whitsunday, and
Mackay/Capricorn management areas (@sfig-depth-map; @sfig-carb-budget-map). Resilient reefs
in the Far Northern management area were exposed to lower mean DHW levels than other reefs
under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however similar spatial patterns were observed
across GCMs for reefs clustered into the high cover cluster (@sfig-gcm-cluster-map). Spatial
findings are not interpreted as accurate predictions of future resilience due to high
uncertainty in input data, particularly depth and connectivity. While similar findings have
been demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014]
regions, depth refugia analyses have often focused on the potential of mesophotic reefs (\>
30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast,
Venegas et al. [-@venegas2019] found no association between depth and the alleviation of
heat stress across a range of shallow reefs in the Pacific.

@@

Current reef restoration and intervention efforts across the GBR, such as deployment of
coral seeding, coral fragment devices and Crown of Thorns Starfish control, target reefs
with sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there
are no explicit justifications for this depth range, intervention depth is likely
constrained by snorkelling and diving deployment and monitoring logistical accessibility,
visual identification of sites from the surface, in combination with tourism return value.
Of the assessed reefs occurring within the 1-10m depth range (1551 reefs), 59% (920) of
reefs are consistently clustered into the Low and Medium cover clusters across the GCMs (at
the GBRMPA management region scale), with many found to be unable to maintain positive
carbonate budgets under projected conditions. PDA revealed a range of depths from 10-15m
where the probability of a reef being clustered into the High cover cluster increased
substantially. Additionally, two-way partial dependence analysis revealed high incoming
connectivity is important for High cluster assignment for reefs at depths of \>10m.
Considering this 10-15m depth range when assessing possible intervention sites may reduce
the exposure of planted and seeded corals to high thermal stress, while still maintaining
effective logistical access to sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. Two-way partial dependence analyses suggest that incoming and outgoing
connectivity contribute positively to high cover cluster allocation for reefs around 15m
deep. However, our results suggest no direct, first order, relationship between weighted
incoming connectivity and the ability of reefs to maintain positive carbonate budgets under
future conditions was found. There has been contrasting evidence [@bode2018; @hock2017]
surrounding whether recent generation connectivity modelling is able to identify reefs that
are important for recovery processes, with findings dependent on the larval connectivity
model used. These investigations were performed using prior generations of larval
connectivity models parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time
periods, without considering the influence of reef depth on resilience. Using updated 1km^2^
resolution larval connectivity across the whole GBR, with a limited temporal sample of
particle flows, it was found that larval connectivity alone may not be important or reef
resilience and recovery, but should be considered in combination with more influential
factors such as depth.

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great 
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines 
    in coral cover from thermal stress events. Understanding drivers behind reef resilience 
    to climate impacts is critical for conservation planning and intervention strategies. A 
    scenario-based population modelling approach was adopted with larval connectivity dynamics, 
    thermal adaptation, and environmental factors to assess coral reef resilience across GBR 
    locations under future climate projections. Using five Global Climate Models (GCMs) 
    spanning five Equilibrium Climate Sensitivities,  reefs were analysed for coral cover 
    time series and their ability of reefs to maintain positive carbonate production budgets 
    under future conditions. Contrary to expectations, larval connectivity patterns did not 
    correlate with increases in maintenance of positive carbonate budgets. Instead, reef depth 
    emerged as the primary predictor of cover-time series resilience and maintenance of positive 
    carbonate budgets, with deeper reefs benefiting from thermal refugia that reduce bleaching 
    impacts. These findings suggest that depth provides critical 
    protection against thermal stress, contrasting current intervention practices that primarily target easily 
    accessible sites in the 1-10m depth range. These results have important implications 
    for reef management, indicating that depth refugia should be prioritized for further 
    consideration in conservation planning alongside connectivity modelling advancements 
    to maximize coral survival under continuing climate change. 
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While reefs experience similar
magnitude DHWs based on SST, the proportion of a reefs' coral population impacted by heat
stress events depends on reef depth. Results of this investigation suggest that for
non-mesophotic reef ecosystems (i.e., \< 30m), depth contributes greatly to reef resilience
and maintenance positive carbonate budgets under future conditions. Movement of shallow
reefs towards negative carbonate budgets in the future may contribute to phase shifts in
reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to maintain
coral driven CaCO~3~ production and accumulation into the future for longer. Spatially these
results indicate that regions of positive carbonate budget maintenance, under ACCESS-ESM1-5,
and high depth levels were concentrated in the Far Northern, offshore Townsville/Whitsunday,
and Mackay/Capricorn management areas (@sfig-depth-map; @sfig-carb-budget-map). Resilient
reefs in the Far Northern management area were exposed to lower mean DHW levels than other
reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however similar spatial patterns were
observed across GCMs for reefs clustered into the high cover cluster
(@sfig-gcm-cluster-map). Spatial findings are not interpreted as accurate predictions of
future resilience due to high uncertainty in input data, particularly depth and
connectivity. While similar findings have been demonstrated in the Pacific [@smith2014;
@baird2018] and Indo-Pacific [@bridge2014] regions, depth refugia analyses have often
focused on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific.

@@

Current reef restoration and intervention efforts across the GBR, such as deployment of
coral seeding, coral fragment devices and Crown of Thorns Starfish control, target reefs
with sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there
are no explicit justifications for this depth range, intervention depth is likely
constrained by snorkelling and diving deployment and monitoring logistical accessibility,
visual identification of sites from the surface, in combination with tourism return value.
Of the assessed reefs occurring within the 1-10m depth range (1551 reefs), 59% (920) of
reefs are consistently clustered into the Low and Medium cover clusters across the GCMs (at
the GBRMPA management region scale), with many found to be unable to maintain positive
carbonate budgets under projected conditions. PDA revealed a range of depths from 10-15m
where the probability of a reef being clustered into the High cover cluster increased
substantially compared to shallower depths. Considering this 10-15m depth range when
assessing possible intervention sites may reduce the exposure of planted and seeded corals
to high thermal stress, while still maintaining effective logistical access to sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. No direct, first order, relationship between weighted incoming connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions was
found. There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes, with findings dependent on the larval connectivity model used. These
investigations were performed using prior generations of larval connectivity models
parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods, without
considering the influence of reef depth on resilience. Using updated 1km^2^ resolution
larval connectivity across the whole GBR, with a limited temporal sample of particle flows,
it was found that larval connectivity alone may not be important or reef resilience and
recovery, but should be considered in combination with more influential factors such as
depth.
